---
title: "8_multivariate_analysis"
output:
  html_document: default
  pdf_document: default
date: "2024-12-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/hisham.shaikh/OneDrive - UGent/Projects/PE477_PE486_Viral_Production")
```

## Importing libraries and setting working directory

```{r libraries, echo=FALSE}
library(tidyverse)
library(ggsci)
library(vegan)
library(ggfortify)
library(psych)
library(corrplot)
library(colorspace)
library(patchwork)
library(cowplot)

```

## Importing depth profiles\
\
Depths profiles were collected onboar R/V Pelagia with sensors mounted on rosette sampler. Parameters collected include -\
1. Salinity\
2. Temperature\
3. Pressure (used to measure Depth)\
4. Density (derived using Gibbs SeaWater equation)\
5. Conductivity\
6. Oxygen\
7. Turbidity\
8. Fluorescence\


```{r importing ctd profiles, echo=FALSE}

ctd_profiles <- utils::read.csv("./results/ctd_profiles/ctd_profiles.csv")

```


I will perform factor analysis to identify latent variables , apply dimension reduction appraoches like PCA and identify clusters to dissect spatio temporal patterns.\
\
## Both cruises combined\
Before factor analysis, I want to make sure its suitability. I will use Kaiser-Meyer-Olkin (KMO) index and Barlett's test from `psych` R package.\

```{r KMO and Barletts tests}
kmo_result <- KMO(ctd_profiles %>% select(-Location_Station_Number, -Location, -Station_Number, -Depth, -Pressure))
bartlett_result <- cortest.bartlett(ctd_profiles %>% select(-Location_Station_Number, -Location, -Station_Number, -Depth, - Pressure))

print(kmo_result)
print(bartlett_result)

```
### *Kaiser-Meyer-Olkin (KMO) Factor Adequacy*\
The Measure of Sampling Adequacy (MSA) of 0.68 in the KMO test is mediocre, but acceptable for factor analysis. \
Salinity has a low MSA of 0.35, shows weak variance with other variables. Consider removing. \
Moderate MSA were observed for temperature, conductivity, and turbidity, whereas high MSA (> 0.7) observed for density, oxygen, and fluorescence, will provide adequate to considerably towards the factor model. \
\

### *Barlett's test of sphericity*
A large chi sqaure value of 198,668 indicates strong correlations among variables. \
Highly significant (p = 0) indicating corraltion matrix is not an identiy matric and factor analysis is appropriate. \

I will exclude Salinity and perform these tests again.\
```{r KMO and Barletts tests without salinity, include=FALSE}
kmo_result_no_salinity <- KMO(ctd_profiles %>% select(-Location_Station_Number, -Location, -Station_Number, -Depth, -Pressure, -Salinity))
bartlett_result_no_salinity <- cortest.bartlett(ctd_profiles %>% select(-Location_Station_Number, -Location, -Station_Number, -Depth, - Pressure, -Salinity))

print(kmo_result_no_salinity)
print(bartlett_result_no_salinity)

```
Removing salinity improves the dataset for factor analysis.\

*Factor Analysis*

```{r factor analysis, echo=TRUE}
fa_result <- fa(ctd_profiles %>% select(-Location_Station_Number, -Location, -Station_Number, -Depth, -Pressure, -Salinity), 
                nfactors = 2, rotate = "varimax", fm = "ml")
print(fa_result)
```
I extracted 2 factors (ML1 and ML2) that together explain 79 % of the variance. \
The Root Mean Square Residual (RMSR) is 0.01 which indicates good fit for the model. The factors explain correlations between variables well. \
The Root Mean Square Error of Estimation (RMSEA) is  0.692 which is high. RMSEA values below < 0.1 are considerable, but given the simplicity of the data it is still useful. \

Explanation of the loadings:\
1. Temperature: Strong negative loading on ML1 and moderate on ML2; high communality (well-explained by both factors); high complexity (contributes to both factors).\
2. Density: Strong positive loading on ML1 and moderate on ML2; high communality; ad high complexity.\
3. Conductivity: same as temperature. \
4. Oxygen: Moderate positive loading on ML1 and ML2; moderate communality (partially explained by the two factors); and high complexity (contributes significantly to both factors).\
5. Turbidity: Moderate loading on ML1 and high on ML2; high communality; and low complexity (primarily explained by ML2).\
6. Fluorescence: Low loadings on both ML1 and ML2; low communality. It also has high uniqueness- much of its variance is independent of these two factors.  High complexity and weakly contributes to both factors.\
\
Temperature, density, conductivity, turbidity have high lowdings and communality inf=dicating that these variables are strngly rerpresented by the two factors. \
Oxygen moderately cntributes to both, and therfore shouuld be retained.\
Fluorescence have low communality and high uniqueness and is not well-represneted by either variables. Might not contribute significant in further analysis. \

ML1 represnts phyical properties - teperature, density, conductivity.\
ML2 represets particlate and biological properties like turidity and oxygen - indicating biological activty. \

I will also exclude fluorescence from further abysis.


##performing PCA
i will use the original variables istead of the latent ones, just to keep the structure intact for easy interpretation. I will exclude fluorescence from further analysis.
```{r pca, echo=T}
pca_data <- ctd_profiles %>%
  select(-Location_Station_Number, -Location, -Station_Number, -Depth, -Pressure, -Salinity, -Fluorescence)

pca_result <- prcomp(pca_data, scale. = TRUE)

summary(pca_result)


pca_var <- data.frame(PC = 1:length(pca_result$sdev),
                      Variance = (pca_result$sdev^2) / sum(pca_result$sdev^2) * 100)

ggplot(pca_var, aes(x = PC, y = Variance)) +
  geom_col() +
  geom_line(aes(group = 1), color = "red") +
  geom_point() +
  labs(title = "Scree Plot: Variance Explained by PCs", x = "Principal Components", y = "Percentage Variance Explained") +
  theme_minimal()
unique_stations <- unique(ctd_profiles$Location_Station_Number)  # Get unique station names
custom_palette <- pal_d3("category20")(length(unique_stations))  # Generate unique colors
color_mapping <- setNames(custom_palette, unique_stations)  # Map each station to a color

autoplot(pca_result, data = ctd_profiles,  color = "Location_Station_Number") +
  theme_minimal() +
  scale_color_manual(values = color_mapping) +
  labs(title = "PCA of CTD Profiles", x = "PC1", y = "PC2")
```

PC1 explains 88.11 % of the variance- captures almost all meaningful pattersn. PC2 adds aother 8.23, bringing the total to 96.33 % \
The datset is effectively two dimensional.\

``` {r pca loadings}
# Scale loadings to a maximum absolute value of 0.1
pca_loadings <- as.data.frame(pca_result$rotation[, 1:2])  # Loadings for PC1 and PC2
pca_loadings$Variable <- rownames(pca_result$rotation)  # Add variable names

# Scale loadings
max_loading <- max(abs(pca_loadings[, 1:2]))  # Maximum absolute loading value
desired_max <- 4  # Desired maximum magnitude
pca_loadings_scaled <- pca_loadings
pca_loadings_scaled[, 1:2] <- pca_loadings_scaled[, 1:2] / max_loading * desired_max

# PCA Scores for plotting
pca_scores <- as.data.frame(pca_result$x[, 1:2])  # PCA scores (PC1 and PC2)
pca_scores$Location_Station_Number <- ctd_profiles$Location_Station_Number
try(pca_scores <- pca_scores %>%
  tidyr::separate(Location_Station_Number, into = c("Location", "Station_Number"), sep = "_", remove = F))


ggplot() +
  # Plot PCA scores
  geom_point(data = pca_scores, aes(x = PC1, y = PC2, color = Location_Station_Number), alpha = 0.7, size = 3) +
  # Add ellipses grouped by Location
  stat_ellipse(data = pca_scores, aes(x = PC1, y = PC2, group = Location, color = Location), 
               linetype = "dashed", size = 1) +
  # Add loadings as arrows
  geom_segment(data = pca_loadings_scaled, aes(x = 0, y = 0, xend = PC1, yend = PC2), 
               arrow = arrow(length = unit(0.2, "cm")), color = "blue", size = 1) +
  # Add variable names to loadings
  geom_text(data = pca_loadings_scaled, aes(x = PC1, y = PC2, label = Variable), 
            color = "blue", vjust = 1, hjust = 1, size = 4) +
  # Labels and theme
  labs(title = "PCA Biplot with Scaled Loadings and Location Ellipses", 
       x = "PC1", y = "PC2") +
  # Apply custom color palette
  scale_color_manual(values = color_mapping) +
  theme_minimal() +
  theme(legend.position = "right")



```

##Performing clustering

```{r clustering, echo=TRUE}
pca_scores <- as.data.frame(pca_result$x[, 1:2])

# Perform k-means clustering (choose k based on elbow method or silhouette score)
set.seed(123)
kmeans_result <- kmeans(pca_scores, centers = 3)  # Adjust centers as needed
pca_scores$Cluster <- as.factor(kmeans_result$cluster)

# Add cluster assignments back to original data
ctd_profiles$Cluster <- pca_scores$Cluster

ggplot(pca_scores, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(title = "Clusters Based on PCA", x = "PC1", y = "PC2") +
  theme_minimal()


```
```{r echo=TRUE}
# Combine PCA plot with color for Location_Station_Number and shape for Cluster
# Generate custom color palette
custom_palette <- pal_d3("category20")(16)  # Generate 16 distinct colors from the D3 palette
# Define a fixed custom palette for Location_Station_Number
custom_palette <- c(
  "PE477_1" = "#1f77b4", "PE477_2" = "#ff7f0e", "PE477_3" = "#2ca02c",
  "PE477_4" = "#d62728", "PE477_5" = "#9467bd", "PE477_6" = "#8c564b",
  "PE486_1" = "#e377c2", "PE486_2" = "#7f7f7f", "PE486_3" = "#bcbd22",
  "PE486_4" = "#17becf", "PE486_5" = "#aec7e8", "PE486_6" = "#ffbb78",
  "PE477_7" = "#98df8a", "PE486_7" = "#ff9896",
   "PE477" = "#0c1844", "PE486" = "#850000"
  # "PE477" = lighten("#0c1844", 0.4), "PE486" = lighten("#850000", 0.4)
)

# Add Cluster column to the PCA scores dataframe
pca_scores <- as.data.frame(pca_result$x[, 1:2])
pca_scores$Cluster <- ctd_profiles$Cluster
pca_scores$Location_Station_Number <- ctd_profiles$Location_Station_Number

# PCA plot with ellipses per Cluster and stroke for points
ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  geom_point(aes(fill = Location_Station_Number, shape = Cluster), color = 'black', alpha = 0.5, size = 3, stroke = 0.3) +  # Add stroke
  stat_ellipse(aes(group = Cluster, linetype = Cluster), level = 0.95, size = 1) +  # Ellipses per Cluster
  scale_color_manual(values = custom_palette) +  # Custom color palette
  labs(title = "PCA: Location_Station_Number and Ellipses per Cluster with Stroke",
       x = "PC1", y = "PC2", color = "Location/Station", shape = "Cluster") +
  theme_minimal() +
  scale_shape_manual(values = c(21, 22, 24)) 
  theme(legend.position = "right")



```
#Hierarchical clustering

```{r}
# Split Location from Location_Station_Number
distance_matrix <- dist(pca_scores, method = "euclidean")  # Adjust method if needed

# Perform hierarchical clustering
hclust_result <- hclust(distance_matrix, method = "ward.D2")  # Use Ward's method

# Cut tree into clusters (choose the number of clusters, e.g., 3)
pca_scores$Cluster <- as.factor(cutree(hclust_result, k = 3))  # Adjust 'k' for number of clusters

# Add cluster assignments back to original data
ctd_profiles$Cluster <- pca_scores$Cluster
ctd_profiles$Location <- sub("_.*", "", ctd_profiles$Location_Station_Number)

# Create a dendrogram object
dendrogram <- as.dendrogram(hclust_result)

# Convert the dendrogram to a ggplot-friendly format
library(ggdendro)
dendro_data <- dendro_data(dendrogram)

# Add geom points to the dendrogram tips colored by Location
ggplot() +
  geom_segment(data = dendro_data$segments,
               aes(x = x, y = y, xend = xend, yend = yend), size = 0.5) +
  geom_point(data = dendro_data$labels,
             aes(x = x, y = 0, color = ctd_profiles$Location), size = 3, alpha = 0.4) +
  labs(title = "Dendrogram with Locations",
       x = "", y = "Height", color = "Location") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "right")


```
```{r}
# need to check the dendrogram for cluster gaian. i think there is some issue with the #assignment of lcusters wrt to stations.
# Create a mapping of cluster assignment from PCA scores
dendro_data$labels$Cluster <- pca_scores$Cluster

# Plot the dendrogram
ggplot() +
  # Add dendrogram branches
  geom_segment(data = dendro_data$segments,
               aes(x = x, y = y, xend = xend, yend = yend), size = 0.5) +
  # Add colored points to the dendrogram tips based on clusters
  geom_point(data = dendro_data$labels,
             aes(x = x, y = 0, color = Cluster), size = 3) +
  # Add rectangles around clusters based on Cluster information
  geom_rect(data = dendro_data$labels,
            aes(xmin = min(x), xmax = max(x), ymin = 0, ymax = max(y)),
            fill = NA, color = "black", linetype = "dashed", size = 0.5) +
  labs(title = "Dendrogram with Clusters Boxed",
       x = "", y = "Height", color = "Cluster") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "right")

```
Cluster 2 splits itself between two seasons. Though we do see seasonal differences. Cluster 

# Differences in variables b


``` {r}
# Load necessary libraries
library(ggplot2)
library(tidyr)

# Reshape the data into long format for plotting
ctd_long <- ctd_profiles %>%
  pivot_longer(
    cols = c(Salinity, Temperature, Density, Conductivity, Oxygen, Turbidity, Fluorescence),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  dplyr::mutate(
    Variable = factor(Variable, levels = c("Salinity", "Temperature", "Density", 
                                           "Conductivity", "Oxygen", "Turbidity", "Fluorescence"))
  )

# Create grouped boxplots
ggplot(ctd_long, aes(x = as.factor(Cluster), y = Value, fill = as.factor(Cluster))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1) +
  #geom_violin()+
  facet_wrap(~Variable, scales = "free_y", nrow = 2) +
  labs(
    title = "Grouped Boxplots of Environmental Variables by Cluster",
    x = "Cluster",
    y = "Value",
    fill = "Cluster"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")

# Create grouped boxplots
ggplot(ctd_long %>% dplyr::filter(Cluster != 2), aes(x = as.factor(Cluster), y = Value, fill = as.factor(Cluster))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1) +
  #geom_violin()+
  facet_wrap(~Variable, scales = "free_y", nrow = 2) +
  labs(
    title = "Grouped Boxplots of Environmental Variables by Cluster without Cluster 2",
    x = "Cluster",
    y = "Value",
    fill = "Cluster"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")

# Create grouped boxplots
ggplot(ctd_long, aes(x = as.factor(Location), y = Value, fill = as.factor(Location))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1) +
  scale_fill_manual(values = c("PE477" = lighten("#0c1844", 0.4), "PE486" = lighten("#850000", 0.2))) +
  #geom_violin()+
  facet_wrap(~Variable, scales = "free_y", nrow = 2) +
  labs(
    title = "Grouped Boxplots of Environmental Variables by Cruise",
    x = "Cluster",
    y = "Value",
    fill = "Cluster"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")

```


# Kruskal Wallis
```{r}
library(dplyr)

# Perform Kruskal-Wallis test for each variable
kruskal_results <- ctd_long %>%
  group_by(Variable) %>%
  summarise(
    Kruskal_Statistic = list(kruskal.test(Value ~ Cluster, data = cur_data())$statistic),
    p_value = list(kruskal.test(Value ~ Cluster, data = cur_data())$p.value)
  ) %>%
  unnest(cols = c(Kruskal_Statistic, p_value))

# Display the results
print(kruskal_results)


library(FSA)
library(dplyr)


# Function to perform Dunn's test after Kruskal-Wallis
perform_posthoc <- function(data) {
  # Perform Kruskal-Wallis test
  kruskal_result <- kruskal.test(Value ~ Cluster, data = data)
  
  # Perform Dunn's test (post-hoc analysis)
  dunn_result <- dunnTest(Value ~ Cluster, data = data, method = "bonferroni")
  
  # Extract results
  posthoc_results <- dunn_result$res
  
  return(posthoc_results)
}

# Apply the function for each variable
posthoc_results <- ctd_long %>%
  group_by(Variable) %>%
  summarise(PostHoc = list(perform_posthoc(cur_data()))) %>%
  unnest(PostHoc)


print(posthoc_results)
```
All combinations are significantly different from each other. Outliers were not removed as they depict ecological variability.\

\
\

Combining this with latitude and longitude to look into spatial patterns. \
## NEED TO RETHINK THIS PART.missing pe486 stations \\

```{r coordinates, eval=FALSE, include=FALSE}
url<- "https://raw.githubusercontent.com/mdhishamshaikh/NorthSea_Microbial_Abundance_Nutrients/main/nj2020_pe477_pe486_bv_abundance_abiotic.csv"
abundance<- readr::read_csv(url)
str(abundance)

#Extracting first entries per Location/Station_Number combinations, as some stations had multiple depths and this information is missing from the fiel on GitHub.
#The first one is the depth for VP assays.
abundance<- abundance %>%
  filter(Location %in% c("PE477", "PE486") )
ctd_profiles_coord <- merge(ctd_profiles, abundance[, c("Location", "Station_Number", "Latitude", "Longitude")], 
                     by = c("Location", "Station_Number"), 
                     all.x = TRUE)

# View the merged data
head(merged_data)
# Ensure rownames from pca_data match original data
ctd_profiles_filtered <- ctd_profiles_coord[rownames(pca_data), ]

# Check dimensions to ensure alignment
nrow(ctd_profiles_filtered)  # Should match nrow(pca_scores)

# Add cluster information from pca_scores
ctd_profiles_filtered$Cluster <- pca_scores$Cluster

# Spatial visualization
for (
  cluster in c(1:3)) {
print(ggplot(ctd_profiles_filtered %>% dplyr::filter(Cluster == cluster,
                                               Depth < 8), aes(x = Longitude, y = Latitude, color = as.factor(Cluster))) +
  geom_point(size = 3, alpha = 0.2, aes(color = Location_Station_Number)) +
  labs(title = "Spatial Visualization of PCA Clusters",
       x = "Longitude", y = "Latitude", color = "Cluster") +
  theme_minimal() +
  theme(legend.position = "right"))
}

```


# Importing Nutrients and micorbial abundance data 3 depths into an RDA.\
```{r micorbial abundances and nutrients}
url<- "https://raw.githubusercontent.com/mdhishamshaikh/NorthSea_Microbial_Abundance_Nutrients/main/nj2020_pe477_pe486_bv_abundance_abiotic.csv"
abundance<- readr::read_csv(url)
str(abundance)

#Extracting first entries per Location/Station_Number combinations, as some stations had multiple depths and this information is missing from the fiel on GitHub.
#The first one is the depth for VP assays.
abundance<- abundance %>%
  dplyr::filter(Location %in% c("PE477", "PE486")) %>%
  dplyr::select(-c(Temperature, Salinity, ends_with("Sample_Name"), Expt_Date, TON))

ctd_profiles_closest_depth <- read.csv("./results/ctd_profiles/ctd_variables_sampling_depths.csv")
ctd_profiles_closest_depth$Depth <- ctd_profiles_closest_depth$Target_Depth

ctd_profiles_abundance_nutrients <- ctd_profiles_closest_depth %>%
  left_join(abundance, by = c("Location", "Station_Number","Depth")) %>%
  group_by(Location, Station_Number, Depth) %>%
  slice_min(row_number(), n = 1) %>%
  ungroup() %>%
  dplyr::select(-c(Measured_depth, Target_Depth, Distance))  %>%
  mutate(
    Nitrate = if_else(Nitrate < 0, 0, Nitrate),
    Nitrite = if_else(Nitrite < 0, 0, Nitrite),
    Phosphate = if_else(Phosphate < 0, 0, Phosphate),
    Silicate = if_else(Silicate < 0, 0, Silicate)
  ) %>%
  na.omit()

write.csv(ctd_profiles_abundance_nutrients, "./results/ctd_profiles_abundance_nutrients_3_depths.csv", row.names = F)

```

I will rerun factor analysis, PCA, and now run RDA with micorbial abudance as a repsons evariable.

```{r}
kmo_result <- KMO(ctd_profiles_abundance_nutrients %>% dplyr::select(-c(Location_Station_Number, Location, Station_Number, Depth, Pressure, Latitude, Longitude)))
bartlett_result <- cortest.bartlett(ctd_profiles_abundance_nutrients %>% select(-c(Location_Station_Number, Location, Station_Number, Depth, Pressure, Latitude, Longitude)))

print(kmo_result)
print(bartlett_result)
```

The error `system is computationally singular` means that the matrix cannot be inverted - multicollineairty and possibly low variance within variables. 

```{r checking for FA assumptions}

# Checking for low variance first.
low_variance <- apply(ctd_profiles_abundance_nutrients %>% 
                      select(-c(Location_Station_Number, Location, Station_Number, Depth, Pressure, Latitude, Longitude)), 
                      2, 
                      var) == 0.1
print(low_variance)
#All FALSE in low variance. 
#No need to remove
# ctd_profiles_abundance_nutrients <- ctd_profiles_abundance_nutrients[, !low_variance]

# Checking correlation

cor_matrix <- cor(ctd_profiles_abundance_nutrients %>% 
                  select(-c(Location_Station_Number, Location, Station_Number, Depth, Pressure, Latitude, Longitude)), 
                  use = "pairwise.complete.obs")

high_corr <- which(abs(cor_matrix) > 0.9 & abs(cor_matrix) < 1, arr.ind = TRUE)
correlated_pairs <- data.frame(
  Var1 = rownames(cor_matrix)[high_corr[, 1]],
  Var2 = colnames(cor_matrix)[high_corr[, 2]],
  Correlation = cor_matrix[high_corr]
)
print(correlated_pairs)
corrplot(cor_matrix, method = "color", tl.cex = 0.8, tl.col = "black")

# Temperature, density, turbidity, and conductivity are highly correlated with each other. Removing conductivity, turbidity, and density.\
# Total_Bacteria, HNA and LNA bacteria are highly correlated. Removing HNA and LNA.\
# V1 and V2 are highly correlated with Total_Viruses. Remove V1, V2, (and V3)\
# 

# Checking again after removing these variables.

kmo_result_low_corr <- KMO(ctd_profiles_abundance_nutrients %>% dplyr::select(-c(Location_Station_Number, Location, Station_Number, Depth, Pressure, Latitude, Longitude, Density, Conductivity, Turbidity, HNA, LNA, V1, V2, V3)))
bartlett_result_low_corr <- cortest.bartlett(ctd_profiles_abundance_nutrients %>% select(-c(Location_Station_Number, Location, Station_Number, Depth, Pressure, Latitude, Longitude, Density, Conductivity, Turbidity, HNA, LNA, V1, V2, V3)))

print(kmo_result_low_corr)
print(bartlett_result_low_corr)
```


```{r }
# fa_result <- fa(ctd_profiles_abundance_nutrients %>% dplyr::select(-c(Location_Station_Number, Location, Station_Number, Depth, Pressure, Latitude, Longitude, Density, Conductivity, Turbidity, HNA, LNA, V1, V2, V3, Salinity)), 
#                 nfactors = 2, rotate = "varimax", fm = "ml")
# print(fa_result)

library(FactoMineR)
library(factoextra)
library(RColorBrewer)

# Perform PCA on the scaled data

pca_data <- ctd_profiles_abundance_nutrients %>% 
                  select(-c(Location_Station_Number, Location, Station_Number, Depth, Pressure, Latitude, Longitude, Density, Conductivity, Turbidity, HNA, LNA, V1, V2, V3))

pca_result <- PCA(pca_data, scale.unit = TRUE, graph = FALSE)

# Extract PCA results for individuals and variables
pca_ind <- as.data.frame(pca_result$ind$coord)  # Individual coordinates
pca_var <- as.data.frame(pca_result$var$coord)  # Variable coordinates

# Add metadata to PCA individual results
pca_ind <- cbind(pca_ind,
                 Location_Station_Number = as.factor(ctd_profiles_abundance_nutrients$Location_Station_Number),
                 Location = as.factor(ctd_profiles_abundance_nutrients$Location),
                 Depth = ctd_profiles_abundance_nutrients$Depth)

# Add midpoints for arrows
pca_var <- pca_var %>%
  mutate(mid_x = Dim.1 * 1.1, mid_y = Dim.2 * 1.1)


# Define a custom color palette with 15 distinct colors
palette_15 <- colorRampPalette(brewer.pal(9, "Set1"))(15)


# Main PCA plot without legend
main_plot <- ggplot() +
  # Add the unit circle
  annotate("path",
           x = cos(seq(0, 2 * pi, length.out = 100)),
           y = sin(seq(0, 2 * pi, length.out = 100)),
           color = "gray", linetype = "dashed") +
  # Add a cross at (0, 0)
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray30", size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray30", size = 1.5) +
  # Add ellipses by Location
  stat_ellipse(data = pca_ind, 
               aes(x = Dim.1, y = Dim.2, group = Location, color = Location, fill = Location), 
               geom = "polygon", alpha = 0.3, linetype = "dashed", size = 1) +
  # Plot individuals with different colors, shapes, and black borders
  geom_point(data = pca_ind, 
             aes(x = Dim.1, y = Dim.2, fill = Location_Station_Number, shape = Location), 
             size = 4) +  # Adjust point size
  # Plot loadings (arrows for variables)
  geom_segment(data = pca_var, 
               aes(x = 0, y = 0, xend = Dim.1 * 1.5, yend = Dim.2 * 1.5), 
               arrow = arrow(length = unit(0.3, "cm")), color = "black") +
  # Add variable names at the midpoint of arrows
  geom_text(data = pca_var, 
            aes(x = mid_x * 1.5, y = mid_y * 1.5, label = rownames(pca_var)),
            vjust = -0.5, hjust = 0.5, fontface = 'bold') +
  # Apply fixed custom palette
  scale_fill_manual(values = custom_palette) +
  scale_color_manual(values = c("PE477" = "#0c1844", "PE486" = "#850000")) +
  scale_shape_manual(values = c("PE477" = 21, "PE486" = 24)) +
  labs(title = "PCA biplot", x = "PC1 (35.19 %)", y = "PC2 (28.68 %)") +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_text(face = 'bold', size = 12),
        axis.text = element_text(size = 10))  # Remove legends from main plot

# Plot to extract shape legend
shape_legend_plot <- ggplot() +
  geom_point(data = pca_ind, 
             aes(x = Dim.1, y = Dim.2, shape = Location), size = 3) +
  scale_shape_manual(values = c("PE477" = 21, "PE486" = 24), name = "Location") +
  theme_void() +  # Remove axis elements
  theme(legend.position = "right",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9))

# Extract shape legend
shape_legend <- cowplot::get_legend(shape_legend_plot)

# Plot to extract fill legend
fill_legend_plot <- ggplot() +
  geom_point(data = pca_ind, 
             aes(x = Dim.1, y = Dim.2, fill = Location_Station_Number), size = 3, shape = 21) +
  scale_fill_manual(values = custom_palette, name = "Location_Station_Number") +
  theme_void() +  # Remove axis elements
  theme(legend.position = "right",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9))

# Extract fill legend
fill_legend <- cowplot::get_legend(fill_legend_plot)


# Combine legends into a single column
combined_legends <- plot_grid(
  shape_legend,
  fill_legend,
  ncol = 1,  # Arrange in a single column
  align = "v",
  rel_heights = c(1, 3)
)

# Combine the main plot and legends
final_plot <- plot_grid(
  main_plot,           # Main PCA plot
  combined_legends,    # Combined legends
  ncol = 2,            # Arrange side by side
  rel_widths = c(4, 1) # Adjust relative widths
)

# Display the final plot
print(final_plot)

# ggsave(
#   plot = final_plot,
#   filename = "./figures/pca_biplot_3depths.svg",
#   device = "svg",
#   dpi = 800,
#   width = 28,
#   height = 22,
#   units = "cm"
# )

```

# Redundancy Anlaysis

```{r}

library(vegan)

# Standardize the response and explanatory variables
response <- ctd_profiles_abundance_nutrients %>% 
            select(Total_Bacteria, Total_Viruses, VBR
                   )

explanatory <- ctd_profiles_abundance_nutrients %>% 
               select(Salinity, Temperature, #Nitrite, 
                      Phosphate, Silicate,
                      Nitrate, Fluorescence, Oxygen)

response_scaled <- scale(response)
explanatory_scaled <- scale(explanatory)
rda_result <- rda(response_scaled ~ ., data = as.data.frame(explanatory_scaled))
summary(rda_result)
anova(rda_result, by = "axis", permutations = 999)
plot(rda_result, scaling = 2, main = "RDA Biplot")

```




```{r}
# Reshape the data into long format for plotting
ctd_long <- ctd_profiles_abundance_nutrients %>%
  pivot_longer(
    cols = c(Salinity, Temperature, Density, Conductivity, Oxygen, Turbidity, Fluorescence, Total_Bacteria, Total_Viruses, VBR),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  mutate(
    Variable = factor(Variable, levels = c("Salinity", "Temperature", "Density", 
                                           "Conductivity", "Oxygen", "Turbidity", "Fluorescence", "Total_Bacteria", "Total_Viruses", "VBR"))
  )

# Create grouped boxplots by Location
ggplot(ctd_long, aes(x = as.factor(Location), y = Value, fill = as.factor(Location))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1) +
 # scale_fill_manual(values = c("PE477" = lighten("#0c1844", 0.4), "PE486" = lighten("#850000", 0.2))) +
  facet_wrap(~Variable, scales = "free_y", nrow = 2) +
  labs(
    title = "Variables by Location",
    x = "Location",
    y = "Value",
    fill = "Location"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "right")
```


